---
title: "RFFTSG v2"
author: "Neil Pettinger"
date: "2025-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Case Study No. 1
# The Barbershop Theory of Emergency Department Delays

We know - from the second case study in the R Foundation Level 1 course - that Emergency Department (ED) four-hour performance is related to the fullness – the crowding – of the ED

But the scatterplot that we drew in RFASS Case Study No. 2 is a bit broad brush, a bit scattergun. It's just 364 dots – each dot is a day – and the fullness of the ED can change a lot in the course of a day. And four-hour performance can also change a lot in the course of day

Therefore I wondered if we could drill down into the granular detail of these 364 days – and the 57,406 ED attendances that took place over the course of those 364 days. Instead of looking at the *average* fullness each day, and the *average* four-hour performance each day, we’ll look instead at the *exact* fullness each time a patient arrived. So we’re going to take 57,406 retrospective fullness snapshots. (Well, actually, we're only going to take 56,100 retrospective fullness snapshots becuase there were occasions when patients arrived at the ED at the exact same moment.) And then we’re going to look at the *exact* eventual ED length of stay of each patient.

```{r message=FALSE, warning=FALSE}

# Step 0 - load the packages ----------------------------------------------

library(readxl)
library(tidyverse)
```

The file we'll be using is a .xlsx file that you've already saved into your data folder. That's why we'll need the {readxl} package. The {tidyverse} package is just a constant for everything we do!

```{r message=FALSE, warning=FALSE}

# Step 1 - import the data ------------------------------------------------

df_ed_duration_extract <-
  read_xlsx("data/01_ed_duration_extract.xlsx")
```

If this code runs correctly, you should end up with 57,406 observations. These are the ED attendances that took place at Anytown General Hospital during the 52-week (364-day) period Monday 29 September 2014 to Sunday 27 September 2015.

The first proper meaningful thing we're going to do is create a dataframe containing all of the unique ED arrival datetimes. I say "unique" advisedly, because a small number of ED attendances arrived at the same minute.

Here's the code that gets us the list of unique arrival times:

```{r message=FALSE, warning=FALSE}

# Step 2 - create a dataframe of unique arrival_datetimes -----------------

df_unique_arrival_times <-
  df_ed_duration_extract |> 
  filter(arrival_datetime >= as.POSIXct('2014-09-29 00:00',
                                            tz = "UTC")) |> 
  group_by(arrival_datetime) |> 
  summarize(no_of_attendances = n()) |> 
  rename(unique_arrival_datetime = arrival_datetime) |> 
  select(unique_arrival_datetime) |> 
  arrange(unique_arrival_datetime)
```

When you run this code chunk, you should end up with a dataframe (called df_unique_arrival_times) that contains 56,100 variables. These are the *unique* arrival times. And you can see that we got there from 57,406 *non*-unique arrival times with the group_by() line.

Having got to a dataframe with 56,100 rows, we then re-named the variable **unique_arrival_datetime** to emphasize what we've done.

OK, the next step is to find out how full of patients the ED was for each of these 56,100 arrival times. Here's teh code for that:

```{r message=FALSE, warning=FALSE}

# Step 3 - find the ED fullness values for each unique arrival time -------

df_ed_fullness_values <-
  df_unique_arrival_times |> 
  left_join(df_ed_duration_extract,
            join_by(unique_arrival_datetime >= arrival_datetime,
                    unique_arrival_datetime < departure_datetime)) |> 
  group_by(unique_arrival_datetime) |> 
  summarize(ed_fullness = n())

```

It's a join() - a left_join() - so we need to be clear about we're doing here.

We're starting with the df_unique_arrival_times dataframe - this is the one we created in Step 2. To visualize the join, think of this as "the table on the left".

The left_join() function introduces "the table on the right". This is the dataframe we imported in Step 1. It contains all of the ED attendances that overlapped the 364-day period. (By 'overlapped', I mean that the dataframe includes patients who arrived *before* Monday 29 September 2014 and patients who departed *after* Sunday 27 September 2015).

Now for the join itself. And it's a non-equi join. In the plainest English I can muster, this is what we are doing: we are asking R to compare the unique_arrival_datetimes from the df_unique_arrival_times dataframe with the arrival_datetimes from teh df_ed_duration extract and to extracyt then when the first is greater than or equal to teh second. 

In other words, if the patient arrived eitehr before or bang on the unique arrival time, iclude it.

But we're not finished becuase there are two parts to teh non-equi join.The second part looks at the departure datetime of each ED attendance and compares it to the unique_arrival_datetimes. When it is greater than the unique arrival time, then it is to be included.

Boyh of the non-equi conditions need to be met in order for the join to hapen.

If we run this code chink at the end of the left_join() line (i.e. before we do the group_by(), we'll get a dataframe with a lot of variables. 1,332,037 to be exact! A varibale for every ED occupied bed minute in the 364 day period. So if there were - on average - 23.7 patients in the departemtn, then that large number is basically 23.7 multiplied by he number of unique arrival times: 23.7 x 56,100 = 1,332,037.

But we tehn go on to group this data to make the dataferam much more manageable.

The next step is to calculate the average klength of stay in the ED for each of the uniquearrval times. Note that in the vast majority of cass the vaues we'll be getting here will be single pateimt vlaus. If the time is 2014-09-29 00:57:00 (the ifrst time in the chronological list) and only one patient arrived at that time, then the length of stay fogure we are lookingfor is that patient's eventual ED length of saty.

But for those (rare) occasions when more than ne patient arried atteh D simultaneously (or in the same minute, at any rate), then we will need to calucalte teh avearge length of saty int eh ED.

```{r message=FALSE, warning=FALSE}

# Step 4 - find the mean ED los_mins values for each unique arrival time --

df_ed_mean_los_mins <-
  df_ed_duration_extract |> 
  filter(arrival_datetime >= as.POSIXct('2014-09-29 00:00',
                                        tz = "UTC")) |> 
  group_by(arrival_datetime) |> 
  summarize(mean_ed_los_mins = mean(los_mins))

```

The next thing we need to do is join the mean_los datafram to the fullness values daatframe.

```{r message=FALSE, warning=FALSE}

# Step 5 - join the ED los_mins values to the ED_fullness values ----------

df_final <-
  df_ed_fullness_values |> 
  left_join(df_ed_mean_los_mins,
            join_by(unique_arrival_datetime == arrival_datetime))

```

This is a straightforward join. Both of the dataframes are of equal length (56,100 rows). We;re simply appening the mean length of stay values to the ED fullnes vlaues.

The tabel we've now got is good enough to create a grpah from.And we will do that. But - be warned! - it will be a very big graph, 56,100 pairs of x, y coordinates! It might take R a few seconds to do this...

```{r message=FALSE, warning=FALSE}

# Step 6 - draw a massive scatterplot -------------------------------------

ggplot(data = df_final) +
  aes(x = ed_fullness,
      y = mean_ed_los_mins) +
  geom_point(alpha = 0.1) +
  scale_y_continuous(limits = c(0, 1440),
                     breaks = seq(0, 1440, 180))

```

There is a limit to hiow useful we can find this scatterplot. There is - actually - just abot - a pattern that's discernble. The fuller the ED, the longer, the length fo saty. But ther eare so many data points that it;sreally just a bit of a mess,

So, to remedy his, we'll go back to df_final and summarize it:

```{r message=FALSE, warning=FALSE}

# Step 7 - summarize the df_final data ------------------------------------

df_final_x <-
  df_final |> 
  group_by(ed_fullness) |> 
  summarize(mean_ed_los_mins = mean(mean_ed_los_mins),
            no_of_arrivals = n())
```

What we've done here is create a summary of df_final (which we've called df_final_x) which gives us - for ewach level of ED fullness that was encoutnered by patienst arriving (from a low value of 1 to a high level of 57)


The final step is to draw a bubbleplot of this summary table:

```{r message=FALSE, warning=FALSE}

# Step 8 - draw a bubbleplot using the summarized data --------------------

ggplot(data = df_final_x) +
  aes(x = ed_fullness,
      y = mean_ed_los_mins,
      size = no_of_arrivals) +
  geom_point(pch = 21,
             fill = "yellow",
             colour = "black",
             alpha = 0.7) +
  geom_smooth(data = df_final_x |> 
              filter(ed_fullness < 60),
                method = lm, 
              se = FALSE, 
              colour ="black",
              linetype = "dotted",
              linewidth = 0.9) +
  scale_x_continuous(limits = c(0, 60),
                     breaks = seq(0, 60, 10)) +
  scale_y_continuous(limits = c(120, 240),
                     breaks = seq(120, 240, 15)) +
  labs(title = "The Barbershop Theory of ED Delays",
       subtitle = "The fuller the ED on arrival, the longer the ED length of stay",
       x = "No. of patients in ED on arrival",
       y = "Length of stay in ED (mins)") +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm"))

```

There's quite a lot of code here, and we won;t implelenmt all of it at once.

## Case Study No. 2
# A day in the life of an Emergency Department

If the first case study helped us understand the relationship between the fullness of an ED and the length fo stay that's lilely to result from a particular level of fullness, then this second case study drills right down to see exactly what ED fullness looks like.

The fullness of an ED varies a lot as we go through the day

